---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
  comment = "#>",
  out.width = "100%"
  )
```

# binsmlr

<!-- badges: start -->
  [![Travis build status](https://travis-ci.com/rrunner/binsmlr.svg?branch=master)](https://travis-ci.com/rrunner/binsmlr)
<!-- badges: end -->

## Overview

binsmlr, as in *bin similar*, is a package to create bins that
are assessed to be heterogenous (in terms of default risk) across a range of
ordinal values. The binning can be performed automatically, interactively and
manually.

The objective is to create groups of data (bins) across a range of ordinal
values. The bins produced should be homogenous within and heterogenous across
bins in terms of the default risk. A binning approach, in the context of data
transformation, is common in different credit risk applications and in 
particular to credit risk portfolios having a relatively large number of
defaults. A prerequisite is that each observation is assigned an ordinal
value such as score or rating etc., that indicates the relative default risk
as proposed by the rank ordering model. Also, a default indicator is
required.

The main process is the following:

1. Create a relatively large set of initial bins.

2. Perform homogeneity tests between all adjacent bins

3. Merge the pair of bins that are the most similar in terms of proportions of 
state of non-default and state of default.

4. Repeat step 2 and 3 until all bins are heterogeneous or if a minimum number 
of bins are obtained.

The process above should result in bins that are statistically different in
terms of default risk across a range of ordinal data.

## Installation

Install the development version from GitHub.

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("rrunner/binsmlr")
```


## Usage

```{r, eval = FALSE, message = FALSE}
library(binsmlr)
help("binsmlr")

# example of interactive binning

# create initial bins
bins <- create_initial_bins(bin_data, 30, "score", "default")
length(bins)
is_monotonic(bins)

# reduce bins by performing repeated homogeneity tests
new_bins <- reduce_bins(bins, min_required_bins = 7, confidence_level = 0.01)
length(new_bins)
is_monotonic(new_bins)

# plot initial and reduced bins
bins_df <- dplyr::bind_rows(bins)
plot(x = bins_df$mid_score, y = log(bins_df$odds), type = "p",
     col = "lightblue", cex = 1.5, pch = 20, ylab = "log(odds)",
     xlab = "score")

new_bins_df <- dplyr::bind_rows(new_bins)
points(x = new_bins_df$mid_score, y = log(new_bins_df$odds),
       col = "darkblue", cex = 1.5, pch = 20)
legend(x = "topright", legend = c("Initial bins", "Reduced bins"),
       col = c("lightblue", "darkblue"), pch = 20, pt.cex = 1.5, bty = "n")
```

## Getting help and report issues

If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/rrunner/binsmlr/issues).
